Dark Control:\\ A Unified Account of Default Mode Function\\ by Control Theory and Reinforcement Learning

### 2017-04-21 paper/paper.tex
2nd half of super pass

\documentclass[10pt,letterpaper]{article}
\usepackage[top=0.85in,left=2.75in,footskip=0.75in]{geometry}

@@

% official NIPS keywords
\textbf{\\keywords}: systems biology, mind wandering, cognitive science,
artificial intelligence,
reinforcement learning
% autonomous learning systems
% \textbf{\\Get opinion from}: Danielle Bassett, Kai Vogeley, Mohammad Shakir,
% Guillaume Dumas, Smallwood, Tenenbaum

@@

\subsection{The prefrontal cortex: stimulus-value association and action consideration}
Analogous to the PMC,
the dorsomedial PFC (dmPFC) of the DMN is believed to subserve
supra-sensory processes
across time, space, and content domains to
exert top-level control on behavior.
Comparing to the PMC, however,
dmPFC function may be closer to a
``mental sketchpad'' \citep{goldman1996prefrontal}, as it
potentially subserves the de-novo generation and manipulation
of meaning representations instructed by stored semantics and memories
\citep{bzdok2013segregation}.

@@

\subsection{The right and left TPJ: prediction error signaling and world semantics}
The DMN emerges with its midline structures early in human development
\citep{doria2010}.
However,
the right and left TPJ may become fully integrated into this major brain
network only after birth.
The TPJs are known to exhibit hemispheric differences
based on cytoarchitectonic borders and gyrification pattern
\citep{seghier2013angular}.
Globally, neuroscientific investigations on hemispheric functional specialization
have highlighted the right versus left cerebral hemisphere as dominant for
attention versus language functions
\citep{seghier2013angular, bzdok2013tpj, bzdok2016left}.

The TPJ in the right-hemispheric DMN (RTPJ)
has been shown to be intimately related to
supra-modal prediction and error signaling.
It is probably central for
action initiation during externally structured tasks and
sensorimotor control by integrating supra-modal stimulus-guided attention
\citep{corbetta2002control}.
Involvement of this DMN node was repeatedly reported in
multi-step action execution \citep{hartmann2005takes},
visuo-proprioceptive conflict \citep{Balslev2005}, and
multi-modal detection of sensory changes across
visual, auditory, or tactile stimulation in a multi-modal fMRI study
\citep{downar2000multimodal}.
In humans, direct electrical stimulation of the
RTPJ during neurosurgery was associated with altered perception
and stimulus awareness \citep{blanke2002neuropsychology}.
%
It was argued that the RTPJ encodes actions and ensuing outcomes
without necessarily relating those to outcome value
\citep{liljeholm2013neural, hamilton2008action,
jakobs2009effects}.
Neural activity in the RTPJ has been argued to be responsible
for stimulus-driven attentional reallocation to
salient and surprising sources of information
as a ‘‘circuit breaker’’ that recalibrates control and maintenance systems
\citep{bzdok2013tpj, corbettashul2008}.
% (Bzdok et al., 2013; Corbetta et al., 2008).
Indeed, patients with RTPJ damage have particular difficulties
with multistep actions \citep{hartmann2005takes}.
In the face of large discrepancies between actual and previously predicted
environmental events the RTPJ acts as a potential switch between
externally-oriented mind sets focussed on the
sensory world and internally-oriented mind sets focussed
on self-relevant mental scene construction.
Transient RTPJ lesion in humans for instance diminished the
impact of predicted intentions of other individuals
\citep{young2010disruption},
a capacity believed to be enabled by the DMN.
The RTPJ might hence be an important relay that shifts away
from the ‘‘internally directed’’ baseline processes
to, instead, deal with (unexpected) environmental objects and events.
% (TODO: slow endogeneous changes in state transitions by the policy and value matrices)

The TPJ in the left-hemispheric hemisphere (LTPJ),
in turn, exhibits a close topographical relationship to Wernicke's area
involved in understanding spoken and written language.
Neurological patients with damage caused to Wernicke's area
have a major impairment of language comprehension
when listening to others or reading a book.
Their speech
preserves natural rhythm and about normal syntax, yet the
voiced sentences are devoid of meaning (i.e., aphasia).
Abstracting from the typical semantic interpretations in linguistics
and neuropsychology,
the LTPJ probably mediates access to and integration of world knowlege,
such as required during action considerations
\citep{binder2011neurobiology, seghier2013angular}.
For instance, LTPJ lesions also entail problems in recognizing
others' pantomimed action towards objects
without obvious relation to processing explicit language content
\citep{varney1987locus}.
% (Varney and Damasio 1987; Rothi et al. 1991; Buxbaum et al. 2005).
%
Inner speech also hinges on knowledge recall
about the physical and inter-personal world.
Indeed,
the internal production of
verbalized thought ("language of the mind") was closely related to the LTPJ
in a multivariate analysis of brain volume
\citep{geva2011neural}.
Further,
episodic memory recall and imagination strongly draw on
reassembling world knowledge.
Isolated building blocks of world statistics probably get reassembled
in internally generated visual scenarios that
navigate present action, weigh hypothetical possibilities, and forecast the future.
%
The LTPJ may hence facilitate the automated prediction of events
by incorporating experience-derived representations of the world
into ongoing action, planning, and problem solving.
% world -> policy matrix driven computation
% self -> value matrix driven computation

\section{Reinforcement learning: a process model for DMN function}
We now argue the outlined neurobiological properties
and potential functions of the DMN nodes
to be sufficient for implementing all components
of a full-fledged RL algorithm.
Recalling past experience, considering candidate actions,
random sampling of possible scenarios, as well as
evaluating instantaneous, delayed, and expected reward outcomes
are key components of intelligent RL agents
that are plausible to intersect in the DMN.

@@

RL is a problem-solving technique in which, through interactions with an unknown \textit{environment / world},
an agent learns to reach goals, optimize reward signals, and minimize costs
in an iterative trial-by-error fashion. See Fig. \ref{fig:rl}.
At a given moment, each taken \textit{action} $a$ triggers an \textit{partially observed} change
in the \textit{state} of the environment
$s \rightarrow s'$, accompanied by environmental feedback signals as \textit{reward}
$r = r(s, a)$ (or \textit{punishment},
since it can be negative) collected by the
agent. This is illustrated in Fig. \ref{fig:rl}.
In this view, the environment is partially controlled by
the action of the agent and the reward can be thought
of as satisfaction -- or punishment -- accompanying the execution of
an action.

Action may be delayed to achieve substantial increases in expected reward
that can grow with time.
The environment is generally taken as stochastic,
that is, changing in random ways. In addition,
the environment is only \textit{partially observed}
in the sense that only part of the current state is observed by
the agent \citep{starkweather2017dopamine}.
The volatility of the environment
is realistic in a computational model which sets out
to explain DMN function.
% XXX: the following passage is already in the introduction
% Indeed, it is known that the more the external world is predictable,
% the more mental activity becomes detached from the actual sensory inputs
% \citep{antrobus1966studies, pope1978regulation},
% the more stimulus-independent thoughts occur, and
% the higher the neural activity in the DMN.
% Conversely, the more the currently executed task is unknown and unpracticed,
% the lower the DMN activity
% \citep{filler1973daydreaming, teasdale1995stimulus}.
% %
% These ``offline'' processes may however contribute to
% optimizing control of the organism.
% Such a framework of DMN funtion based on reinforcement learning
% can naturally embed human behavior
% in the tension between exploitative action with immediate gains and
% explorative action with longer-term reward schedules
% \citep{dayan2008decision}.
We argue that a functional account of the DMN based on RL
can naturally embed human behavior
in the tension between exploitative action with immediate gains and
explorative action with longer-term reward outcomes
\citep{dayan2008decision}.
DMN implication in a diversity of
particularly advanced cognitive processes
can be parsimoniously explained as probabilistic mental
scene simulations coupled with prediction error minimization
to calibrate action trajectories and
maximize reward outcomes at different time scales.
Such a purposeful optimization objective
may be solved by a stochastic approximation
based on a brain implementation of MCMC sampling
\citep{tenenbaum2011grow}.

\subsection{Partially Observable Markov Decision Processes}
We argue that the brain in general and the DMN in particular
are a physical system governed by the laws of
physics and can be locally described by Markov processes at a sufficiently coarse scale.
Indeed, as noted in \citep{tegmark2016improved},
any system obeying the laws of classical physics can be accurately modeled as a Markov process as long as the time
step is sufficiently short.
% , defining $s(t)$, the state at time $t$, as the position in state space.
The process has $memory$ if the next state depends not only on the current state but also on some finite
number of past states.
Rational probabilistic planning can be reformulated
as a standard memoryless Markov process by simply expanding the
definition of the state $s$ to include elements of the past.

In artificial intelligence and machine learning, a popular computational model for
multi-step decision processes in such an environment are
\textit{Partially Observable Markov Decision Processes (POMDPs)}\citep{sutton1998reinforcement}.
A POMDP formalizes a sequential decision process in which it is assumed that environment dynamics are determined by a Markov process,
but the agent cannot directly observe the underlying state.
Instead, the agent tries to optimize a \textit{subjective} reward
signal (i.e., is likely to be different for another agent in the same state).
This is performed alongside each transition in the environments state space, and maintains a probability distribution over the set of possible
states, based on a set of observations and observation probabilities. This is a minimal amount of assumptions that can be made about an environment,
and is characteristic of so-called \textit{model-free RL},
which we propose as a useful formal framework for DMN function.
%

\paragraph{Model-free Reinforcement Learning.}
Model-free RL has had great success in real-world problems including robotics\citep{abbeel2004,andrewng}, computer game play, \citep{mnih2015,silver2016mastering},
etc., just to name a few. Such a learning scheme is plausibly realizable in the human brain. Indeed,
a core property of human intelligence has been proposed underly to improve
on an expected utility outcomes as a strategy for action choice in uncertain
environments \citep{gershman2015computational}, a situation perfectly captured by the formalism of POMDPs. Also,
it has long been proposed \citep{dayan2008decision} that there
is a rather direct mapping of model-free RL learning algorithms
onto aspects of the brain.
The neurotransmitter dopamine could serve
as a ``teaching signal'' to better estimate value associations
and action policies by controlling
synaptic plasticity at targets such as the ventral and
dorsolateral striatum.
In contrast, \textit{model-based RL} would start off with some mechanistic assumptions about the dynamics of the world.
These can be assumptions about the physical laws governing the agent's world or constraints on the state space and transitions between states.
For instance, if a rat in a maze knows
that standing still will produce no change in the environment, and in particular will not eventually lead to finding the food.
%

@@

\subsection{The knots and bolts of Reinforcement Learning in the DMN}
\textbf{XXX Elvis: ==> Elvis a short preface is needed here, explaining the roadmap of this subsection from a high-level perspective!}
\subsubsection{The objective value of an action given a state}
Given only the limited information available from a POMDP, at a state $s$ the average quality or usefulness of taking an action $a$ under a policy $\pi$ is completely captured by the single number
% We can now formulate the purpose of the DMN
% as finding a policy $\pi$ for the human agent that maximizes the
% expected cumulative value of an state-action pair $(a,s)$, also known as the \textit{Q-value}, namely
\begin{equation}
  \label{eq:qval}
  Q^{\pi}(s,a) = \mathbb E [G^\pi|s_0=s,a_0=a],
\end{equation}
called the $Q$-value for the state-action pair $(s,a)$.
In other words, $Q^{\pi}(s,a)$ corresponds to the expected reward
over all possible action trajectories, in which
the agent sets out in the environment in state
$s$, chooses action $a$, and then follows the policy $\pi$ to select future actions.
%% Given that each $\pi$ correspondence to an exhaustive description of
%% behavioral patterns, $Q^{\pi}(s_t,a_t)$ corresponds to the epistemic value of
%% the set of actions given states in finite time.
For the brain,
$Q^{\pi}(s, a)$ defined in \eqref{eq:qval} provides the subjective
utility of executing a specific action; it answers the question
"What is the expected utility of taking action $a$ in this situation ?".
$Q^{\pi}(s,a)$ offers a formalization of optimal behavior that
may well reflect processing aspects of the DMN in human agents.

@@

The goal of learning is to approach the policy $\pi^*$ as ``close as possible'', that is to \text{solve} the POMDP.
Of course \eqref{eq:qlearning} is just a definition and
represents by no means, a candidate schema for solving POMDPs with even moderately-sized action and state spaces: it's intractable!
% where $\tilde{Q}(s, a)$ is the current estimate of the state-action value
% table.
Fortunately, the \textit{Bellman equation} \citep{sutton1998reinforcement} provides a fixed-point relation which defines $Q^*$ implicitly via a sampling procedure, without querying the entire space of policies, viz
\begin{equation}
  Q^* = \text{Bel}(Q^*),
  \label{eq:bellman}
\end{equation}
where the ``Bellman transform'' $\text{Bel}(Q)$ of an arbitrary $Q$-value function $Q: \mathcal S \times \mathcal A \rightarrow \mathbb R$  is defined by setting
\begin{equation}
  \begin{split}
   \text{Bel}(Q)(s,a) &:=
   \mathbb E_{s' \sim p(s'|s,a)} [r(s,a) + \gamma \max_{a' \in \mathcal A}Q(s', a')]\\
   &= r(s,a) + \gamma\mathbb E_{s' \sim p(s'|s,a)} [\max_{a' \in \mathcal A}Q(s', a')]\\
   &= \text{instantaneous reward} + \text{expected reward for acting greedily thereafter}
%    \; \forall (s,a) \in \mathcal S \times \mathcal A
    % \\&=
    % \mathbb E_{s' \sim p(s'|s,a)} [r(s,a) + \gamma Q^*(s', \pi^*(s'))].
    \end{split}
  \end{equation}
The Bellman equation \eqref{eq:bellman} is a ``self-consistency'' equation which provides
a recursive decomposition of optimal behavior by dividing the $Q$-value function into the immediate
reward and the discounted rewards of the upcoming states.
% The optimal Q-value operator $Q^*$
% is a fixed point for this equation.

@@

As a consequence, the complicated dynamic programming optimization problem \eqref{eq:qlearning}
is broken into simpler sub-problems at different time points.
Exploitation of hierarchical structure in actions consideration
has previously been closely related to the medial prefrontal part of the DMN
\citep{koechlin1999role, braver2002role}.
Using the Bellman equation, each state can be associated with a certain value
to guide action towards a better state, thus improving on the agent's current policy.
% It is a fixed-point equation for the deterministic policy $\pi$.
Note that in \eqref{eq:bellman} the sampling is done only over things which
depend on the environment, and so can be learned off-policy by observing state transitions
triggered by another behavioral policy, which can be stochastic.

@@

\begin{equation}
  \pi^{\text{greedy}}(s) = \argmax_{a \in \mathcal A}\tilde{Q}(s, a|\theta),
  \label{eq:policy}
\end{equation}
by querying an approximate version $a \mapsto \tilde{Q}(a,s|\theta)$ of the $Q$-value function ``frozen'' at state $s$. For instance, a simplistic linear model with a kernel $\phi$ would be of the
form $\tilde{Q}(s, a|\theta) = \phi(s,a)^T\theta$, where
$\phi(s,a)$ would represent a high-level representation of the state-action pairs
$(s,a)$, as proposed by \citep{songNIPS2016} for example.
%  and Parr et al. 2003 (JMLR)].
In the DMN, The dmPFC (BA9) would implement such a greedy lookup over the action space.
The tensor $\theta$ encodes the model
  parameters and would correspond to synaptic weights and / or connection strengths between
  DMN regions. It is a time-varying neuronal program which dictates how to move from world states $s$ to actions $a$ via the greedy policy \eqref{eq:policy}.
  The oracle Q-values $Q(s, a|\theta)$ tell the DMN the (expected) usefulness of taking an action a in state / situation  $s$. The DMN (via BA9 ?) would then choose, at a given $s$, an action $a$ which maximizes this tabulated values. This mapping from states to actions defines a matrix, which can be conveniently called a "policy matrix". Learning consists in starting from a given instantiation of such a table and tweaking it around while using it to take actions which take the agent to different areas of the table, and so on, and so forth.

  \subsubsection{Self-training and the loss function}
There can be no learning without a loss function in sight.
The action $a$ taken in state $s$ according to the policy matrix defined in
\eqref{eq:policy} triggers a reward $r$ collected by the agent,
% (measured on their internal currency of worldly values ?),
after which the environment transitions to a new state $s' \in \mathcal S$.
% \textbf{XXX Please try to smoothen the previous sentence}
At the end of this cycle, a new \textit{experience} $e = (s,a,r,s')$ is produced, which represents an exemplar
behavior of the agent and is recorded in replay memory buffer --which we hypothesize to be the
hippocampus--, possibly discarding the oldest entries to make space:
$\mathcal D \leftarrow \text{append}(\mathcal D, e)$.
At time step $k$, the agent seeks an update $\theta_{k} \leftarrow \theta_{k-1} + \delta \theta_{k}$ of the parameters for its approximate model of $Q$-value functional. This warrants learning,
and a loss function is in demand.
% \subsubsection{Stochastic gradient in the DMN via experience replay}
The Bellman equation \eqref{eq:bellman} provides a way to produce such a loss function \eqref{eq:oracle} as we now outline.
% A practically successful RL approach
% needs ways to efficiently approximate the state-action value function
% to allow for scaling to large state and action spaces.
Replay consists in sampling (uniform or importance-weighted
\footnote{e.g weighted by TD error of the state transition $s \overset{a}{\rightarrow} s'$.} ?)  mini-batches of experiences
$(s, a, r, s') \sim \mathcal D$ from the replay memory $\mathcal D$.
The agent then tries to approximate
the would-be $Q$-value for the state-action pair $(s,a)$ as predicted by the Bellman equation \eqref{eq:bellman}, namely
\begin{equation}
  y_k := y_k(s,a,s') =  r + \gamma \max_{a'} \tilde{Q}(s', a'|\theta_{k-1}),
\end{equation}
with the prediction of a parametrized regression model $(s,a)
\mapsto \tilde{Q}(s, a|\theta_{k-1})$.
From a neurobiological perspective,
experience replay would be manifested as the re-occurrence of a
neuron spiking sequence that also occurs during specific actions.
%, but the replay has a much faster time scale.
In the human brain, the hippocampus is likely to contribute to such a mechanism,
as neuroscience experiments on various animals have repeatedly indicated
in rats, mice, cats, rabbits, songbirds, and
monkeys \citep{buhry2011,nokia2010,dave2000,skaggs2007}.

>>>>

sensory\documentclass[10pt,letterpaper]{article}
\usepackage[top=0.85in,left=2.75in,footskip=0.75in]{geometry}

@@

% official NIPS keywords
\textbf{\\keywords}: systems biology, mind-wandering, cognitive science,
artificial intelligence,
reinforcement learning
% autonomous learning systems
% \textbf{\\Get opinion from}: Danielle Bassett, Kai Vogeley, Mohammad Shakir,
% Guillaume Dumas, Smallwood, Tenenbaum

@@

\subsection{The prefrontal cortex: stimulus-value association and action consideration}
Analogous to the PMC,
the dorsomedial PFC (dmPFC) of the DMN is believed to subserve
multi-sensory processes
across time, space, and content domains to
exert top-level control on behavior.
Comparing to the PMC, however,
dmPFC function may be closer to a
``mental sketchpad'' \citep{goldman1996prefrontal}, as it
potentially subserves the de-novo generation and manipulation
of meaning representations instructed by stored semantics and memories
\citep{bzdok2013segregation}.

@@

\subsection{The right and left TPJ: prediction error signaling and world semantics}
The DMN emerges with its midline structures early in human development
\citep{doria2010}.
However,
the right and left TPJ may become fully integrated into this major brain
network only after birth.
The TPJs are known to exhibit hemispheric differences
based on cytoarchitectonic borders and gyrification pattern
\citep{seghier2013angular}.
Globally, neuroscientific investigations on hemispheric functional specialization
have highlighted the right versus left cerebral hemisphere as dominant for
attentional versus semantic functions
\citep{seghier2013angular, bzdok2013tpj, bzdok2016left}.

The TPJ in the right-hemispheric DMN (RTPJ)
has been shown to be intimately related to
multi-sensory prediction and error signaling.
It is probably central for
action initiation during goal-dircted psychological tasks and for
sensorimotor control by integrating multi-sensory attention
\citep{corbetta2002control}.
Involvement of this DMN node was repeatedly reported in
multi-step action execution \citep{hartmann2005takes},
visuo-proprioceptive conflict \citep{Balslev2005}, and
multi-sensory detection of environmental changes across
visual, auditory, or tactile stimulation in an fMRI experiment
\citep{downar2000multimodal}.
In humans, direct electrical stimulation of the
RTPJ during neurosurgery was associated with altered perception
and stimulus awareness \citep{blanke2002neuropsychology}.
%
It was argued that the RTPJ encodes actions and ensuing outcomes
without necessarily relating those to value estimation
\citep{liljeholm2013neural, hamilton2008action,
jakobs2009effects}.
Neural activity in the RTPJ has been argued to reflect
stimulus-driven attentional reallocation to
self-relevant and unexpected sources of information
as a ‘‘circuit breaker’’ that recalibrates control and maintenance systems
\citep{bzdok2013tpj, corbettashul2008}.
% (Bzdok et al., 2013; Corbetta et al., 2008).
Indeed, patients with RTPJ damage have particular difficulties
with multistep actions \citep{hartmann2005takes}.
In the face of large discrepancies between actual and previously predicted
environmental events the RTPJ acts as a potential switch between
externally-oriented mind sets focussed on the
sensory environment and internally-oriented mind sets focussed
on self-relevant mental scene construction.
For instance, temporally induced RTPJ lesion in humans diminished the
impact of predicted intentions of other individuals
\citep{young2010disruption},
a capacity believed to be enabled by the DMN.
The RTPJ might hence be an important relay that shifts away
from the ‘‘internally directed’’ baseline processes
to, instead, deal with (unexpected) environmental objects and events.
% (TODO: slow endogeneous changes in state transitions by the policy and value matrices)

The TPJ in the left-hemispheric hemisphere (LTPJ),
in turn, exhibits a close relationship to Wernicke's area
involved in understanding semantics, such as in spoken and written language.
Neurological patients with damage in Wernicke's area
have a major impairment of language comprehension
when listening to others or reading a book.
Patient speech
preserves natural rhythm and normal syntax, yet the
voiced sentences lack meaning (i.e., aphasia).
Abstracting from the typical speech interpretations in linguistics
and neuropsychology,
the LTPJ probably mediates access to and integration of world knowlege,
such as required during action considerations
\citep{binder2011neurobiology, seghier2013angular}.
For instance, LTPJ lesions in humans also entail problems in recognizing
others' pantomimed action towards objects
without obvious relation to processing explicit language content
\citep{varney1987locus}.
% (Varney and Damasio 1987; Rothi et al. 1991; Buxbaum et al. 2005).
%
Inner speech also hinges on knowledge recall
about the physical and social world.
Indeed,
the internal generation of
verbalized thought ("language of the mind") was closely related to the LTPJ
in a pattern analysis of brain volume
\citep{geva2011neural}.
Further,
episodic memory recall and imagination strongly draw on
reassembling world knowledge.
Isolated building blocks of world statistics probably get reassembled
in internally generated mental scenarios that
guide present action choice,
weigh hypothetical possibilities, and forecast the future.
%
The LTPJ may hence facilitate the automated prediction of events
by incorporating experience-derived building blocks of world statistics
into ongoing action, planning, and problem solving.
% world -> policy matrix driven computation
% self -> value matrix driven computation

\section{Reinforcement learning: a process model for DMN function}
We now argue the outlined neurobiological properties
and functional involvements of the DMN nodes
to be sufficient for implementing all components
of a full-fledged RL system.
Recalling past experience, considering candidate actions,
random sampling of possible experiences, as well as
evaluating instantaneous, delayed, and expected reward outcomes
are key components of intelligent RL agents
that are plausible to intersect in the DMN.

@@

RL is a problem-solving technique in which, through interactions with an unknown \textit{environment},
an agent learns to reach goals, optimize reward signals, and minimize costs
in an iterative trial-by-error fashion. See Fig. \ref{fig:rl}.
At a given moment, each taken \textit{action} $a$ triggers an \textit{partially observed} change
in the \textit{state} of the environment
$s \rightarrow s'$, accompanied by environmental feedback signals as \textit{reward}
$r = r(s, a)$ collected by the agent (Fig. \ref{fig:rl}).
If the collected reward outcome yields a negative value it can be
more naturally interpreted as \textit{punishment}.
In this view, the environment is partially controlled by
the action of the agent and the reward can be thought
of as satisfaction -- or aversion -- accompanying the execution of
an action.

Action may be delayed to achieve substantial increases in expected reward
that is expected to grow with time.
The environment is generally taken as \textit{stochastic},
that is, changing in random ways. In addition,
the environment is only \textit{partially observed}
in the sense that only part of the current state is observed by
the agent \citep{starkweather2017dopamine}.
Assuming volatility of the environment
is realistic in a computational model which sets out
to explain DMN function.
% XXX: the following passage is already in the introduction
% Indeed, it is known that the more the external world is predictable,
% the more mental activity becomes detached from the actual sensory inputs
% \citep{antrobus1966studies, pope1978regulation},
% the more stimulus-independent thoughts occur, and
% the higher the neural activity in the DMN.
% Conversely, the more the currently executed task is unknown and unpracticed,
% the lower the DMN activity
% \citep{filler1973daydreaming, teasdale1995stimulus}.
% %
% These ``offline'' processes may however contribute to
% optimizing control of the organism.
% Such a framework of DMN funtion based on reinforcement learning
% can naturally embed human behavior
% in the tension between exploitative action with immediate gains and
% explorative action with longer-term reward schedules
% \citep{dayan2008decision}.
We argue that a functional account of the DMN based on RL
can naturally embed human behavior
in the tension between exploitative action with immediate gains and
explorative action with longer-term reward outcomes
\citep{dayan2008decision}.

In short, DMN implication in a diversity of
particularly advanced cognitive processes
can be parsimoniously explained as probabilistic mental
scene simulations coupled with prediction error minimization
to calibrate action trajectories to
maximize reward outcomes at different time scales.
Such a purposeful optimization objective
may be solved by a stochastic approximation
based on a brain implementation of MCMC sampling
\citep{tenenbaum2011grow}.

\subsection{Partially Observable Markov Decision Processes}
We argue that the brain in general and the DMN in particular
are a physical system governed by the laws of
physics and can be formally described
by Markov processes at a sufficiently coarse scale.
It has indeed been previously proposed \citep{tegmark2016improved} that
any system obeying the laws of classical physics can be accurately modeled as a Markov process as long as the time
step is sufficiently short.
% , defining $s(t)$, the state at time $t$, as the position in state space.
The process has $memory$ if the next state depends not only on the current state but also on some finite
number of past states.
Rational probabilistic planning can be reformulated
as a standard memoryless Markov process by simply expanding the
definition of the state $s$ to include experience episodes of the past.

In artificial intelligence and machine learning, a popular computational model for
multi-step decision processes in such an environment are
\textit{Partially Observable Markov Decision Processes (POMDPs)}\citep{sutton1998reinforcement}.
A POMDP operationalizes a sequential decision process
in which it is assumed that environment dynamics are determined by a Markov process,
but the agent cannot directly observe the underlying state.
Instead, the agent tries to optimize a \textit{subjective} reward
signal (i.e., is likely to be different for another agent in the same state).
This is performed alongside each transition in the environments state space, and maintains a probability distribution over the set of possible
states, based on a set of observations and observation probabilities. This is a minimal amount of assumptions that can be made about an environment,
and is characteristic of so-called \textit{model-free RL},
which we propose as a useful formal framework for DMN function.
%

\paragraph{Model-free Reinforcement Learning.}
Model-free RL has had great success in real-world problems including robotics\citep{abbeel2004,andrewng}, computer game play, \citep{mnih2015,silver2016mastering},
etc., just to name a few. Such a learning scheme can be
plausibly realized in the human brain. Indeed,
a core property of human intelligence has been proposed underly to improve
on an expected utility outcomes as a strategy for action choice in uncertain
environments \citep{gershman2015computational}, a situation perfectly captured by the formalism of POMDPs.
It has also long been proposed \citep{dayan2008decision} that there
is a rather direct mapping of model-free RL learning algorithms
onto aspects of the brain.
The neurotransmitter dopamine could serve
as a ``teaching signal'' to better estimate value associations
and action policies by controlling
synaptic plasticity at targets such as the ventral striatum.
In contrast, \textit{model-based RL} would start off with some mechanistic assumptions about the dynamics of the world.
These can be assumptions about the physical laws governing the agent's world or constraints on the state space and transitions between states.
For instance, if a rat in a maze knows
that standing still will produce no change in the environment, and in particular will not eventually lead to finding the food.
%

@@

\subsection{The nuts and bolts of Reinforcement Learning in the DMN}
\textbf{XXX Elvis: ==> Elvis a short preface is needed here, explaining the roadmap of this subsection from a high-level perspective!}
\subsubsection{The objective value of an action given a state}
Given only the limited information available from a POMDP, at a state $s$ the average quality or usefulness of taking an action $a$ under a policy $\pi$ is completely captured by the single number
% We can now formulate the purpose of the DMN
% as finding a policy $\pi$ for the human agent that maximizes the
% expected cumulative value of an state-action pair $(a,s)$, also known as the \textit{Q-value}, namely
\begin{equation}
  \label{eq:qval}
  Q^{\pi}(s,a) = \mathbb E [G^\pi|s_0=s,a_0=a],
\end{equation}
called the $Q$-value for the state-action pair $(s,a)$.
In other words, $Q^{\pi}(s,a)$ corresponds to the expected reward
over all considered action trajectories, in which
the agent sets out in the environment in state
$s$, chooses action $a$, and then follows the policy $\pi$ to select future actions.
%% Given that each $\pi$ correspondence to an exhaustive description of
%% behavioral patterns, $Q^{\pi}(s_t,a_t)$ corresponds to the epistemic value of
%% the set of actions given states in finite time.
For the brain,
$Q^{\pi}(s, a)$ defined in \eqref{eq:qval} provides the subjective
utility of executing a specific action; it answers the question
"What is the expected utility of taking action $a$ in this situation ?".
$Q^{\pi}(s,a)$ offers a formalization of optimal behavior that
may well reflect processing aspects of the DMN in human agents.

@@

The learning goal is to approach the
policy $\pi^*$ as ``close as possible'', that is to \text{solve} the POMDP.
Of course \eqref{eq:qlearning} is just a definition and
represents by no means, a candidate schema for solving POMDPs with even moderately-sized action and state spaces: it's intractable!
% where $\tilde{Q}(s, a)$ is the current estimate of the state-action value
% table.
Fortunately, the \textit{Bellman equation} \citep{sutton1998reinforcement} provides a fixed-point relation which defines $Q^*$ implicitly via a sampling procedure, without querying the entire space of policies, viz
\begin{equation}
  Q^* = \text{Bel}(Q^*),
  \label{eq:bellman}
\end{equation}
where the ``Bellman transform'' $\text{Bel}(Q)$ of an arbitrary $Q$-value function $Q: \mathcal S \times \mathcal A \rightarrow \mathbb R$  is defined by setting
\begin{equation}
  \begin{split}
   \text{Bel}(Q)(s,a) &:=
   \mathbb E_{s' \sim p(s'|s,a)} [r(s,a) + \gamma \max_{a' \in \mathcal A}Q(s', a')]\\
   &= r(s,a) + \gamma\mathbb E_{s' \sim p(s'|s,a)} [\max_{a' \in \mathcal A}Q(s', a')]\\
   &= \text{instantaneous reward} + \text{expected reward for acting greedily thereafter}
%    \; \forall (s,a) \in \mathcal S \times \mathcal A
    % \\&=
    % \mathbb E_{s' \sim p(s'|s,a)} [r(s,a) + \gamma Q^*(s', \pi^*(s'))].
    \end{split}
  \end{equation}
The Bellman equation \eqref{eq:bellman} is a ``self-consistency'' equation which provides
a recursive decomposition of optimal behavior by dividing the $Q$-value function into the immediate
reward and the discounted rewards of the upcoming states.
% The optimal Q-value operator $Q^*$
% is a fixed point for this equation.

@@

As a consequence, the complicated dynamic programming optimization problem \eqref{eq:qlearning}
is broken into simpler sub-problems at different time points.
Exploitation of hierarchical structure in action considerations
has previously been closely related to the medial prefrontal part of the DMN
\citep{koechlin1999role, braver2002role}.
Using the Bellman equation, each state can be associated with a certain value
to guide action towards a better state, thus improving on the agent's current policy.
% It is a fixed-point equation for the deterministic policy $\pi$.
Note that in \eqref{eq:bellman} the sampling is done only over things which
depend on the environment, and so can be learned off-policy by observing state transitions
triggered by another behavioral policy, which can be stochastic.

@@

\begin{equation}
  \pi^{\text{greedy}}(s) = \argmax_{a \in \mathcal A}\tilde{Q}(s, a|\theta),
  \label{eq:policy}
\end{equation}
by querying an approximate version $a \mapsto \tilde{Q}(a,s|\theta)$ of the $Q$-value function ``frozen'' at state $s$. For instance, a simplistic linear model with a kernel $\phi$ would be of the
form $\tilde{Q}(s, a|\theta) = \phi(s,a)^T\theta$, where
$\phi(s,a)$ would represent a high-level representation of the state-action pairs
$(s,a)$, as proposed by \citep{songNIPS2016} for example.
%  and Parr et al. 2003 (JMLR)].
In the DMN, The dmPFC (BA9) would implement such a greedy lookup over the action space.
The tensor $\theta$ encodes the model
  parameters and would correspond to synaptic weights and / or connection strengths between
  DMN regions. It is a time-varying neuronal program which dictates how to move from world states $s$ to actions $a$ via the greedy policy \eqref{eq:policy}.
  The oracle Q-values $Q(s, a|\theta)$ tell the DMN the (expected) usefulness of taking an action a in state / situation  $s$.
  The DMN, and in particular its dmPFC node, would then contribute to the choice, at a given $s$, of an action $a$ which maximizes this tabulated values. This mapping from states to actions defines a matrix, which can be conveniently called a "policy matrix". Learning consists in starting from a given instantiation of such a table and tweaking it around while using it to take actions which take the agent to different areas of the table, and so forth.

  \subsubsection{Self-training and the loss function}
Successful learning is challenging without a clear learning goal --
the \textit{loss function}.
The action $a$ chosen in state $s$ according to the policy matrix defined in
\eqref{eq:policy} yields a reward $r$ collected by the agent,
% (measured on their internal currency of worldly values ?),
after which the environment transitions to a new state $s' \in \mathcal S$.
% \textbf{XXX Please try to smoothen the previous sentence}
At the end of this cycle, a new \textit{experience} $e = (s,a,r,s')$ is produced, which represents an exemplar
behavior of the agent and is recorded in replay memory buffer --which we hypothesize to be the
hippocampus--, possibly discarding the oldest entries to make space:
$\mathcal D \leftarrow \text{append}(\mathcal D, e)$.
At time step $k$, the agent seeks an update $\theta_{k} \leftarrow \theta_{k-1} + \delta \theta_{k}$ of the parameters for its approximate model of $Q$-value functional. This warrants learning,
and a loss function is in demand.
% \subsubsection{Stochastic gradient in the DMN via experience replay}
The Bellman equation \eqref{eq:bellman} provides a way to produce such a loss function \eqref{eq:oracle} as we now outline.
% A practically successful RL approach
% needs ways to efficiently approximate the state-action value function
% to allow for scaling to large state and action spaces.
Replay consists in sampling (uniform or importance-weighted
\footnote{e.g weighted by TD error of the state transition $s \overset{a}{\rightarrow} s'$.} ?)  mini-batches of experiences
$(s, a, r, s') \sim \mathcal D$ from the replay memory $\mathcal D$.
The agent then tries to approximate
the would-be $Q$-value for the state-action pair $(s,a)$ as predicted by the Bellman equation \eqref{eq:bellman}, namely
\begin{equation}
  y_k := y_k(s,a,s') =  r + \gamma \max_{a'} \tilde{Q}(s', a'|\theta_{k-1}),
\end{equation}
with the prediction of a parametrized regression model $(s,a)
\mapsto \tilde{Q}(s, a|\theta_{k-1})$.
From a neurobiological perspective,
experience replay would be manifested as the re-occurrence of a
neuron spiking sequence that also occurs during specific actions.
%, but the replay has a much faster time scale.
In the human brain, the hippocampus is likely to contribute to such a mechanism,
as neuroscience experiments on various animals have repeatedly indicated
in rats, mice, cats, rabbits, songbirds, and
monkeys \citep{buhry2011,nokia2010,dave2000,skaggs2007}.
